Co-evolution of lexical meaning \& pragmatic use

### 2018-03-13 paper/cogsci-submission2.tex
thorough revision first half; appendix added; clarified effect of l>1

\begin{abstract}
 According to standard linguistic theory, the meaning of an utterance is the product of
  conventional semantic meaning and general pragmatic rules on language use. To investigate how
  cultural evolution of language plays out under this picture of the semantics-pragmatics
  interface, we present a game-theoretic model of the competition between types of language
  users, each endowed with a selection of lexical representations and a particular pragmatic
  disposition to act on them. Our model traces two evolutionary forces and their interaction:
  (i) pressure towards communicative efficiency and (ii) transmission perturbations during the acquisition of
  linguistic knowledge. We illustrate the model based on a case study on scalar
  implicatures, which suggests the relationship between underspecified semantics and pragmatic inference to be one of co-evolution. The maintenance of underspecified semantics depends on pragmatic reasoning for their enrichment. In the opposite direction, complex semantics that can do away with mutual reasoning may not foster pragmatic reasoning, and may even be encumbered by it.
\end{abstract}

@@

A particularly productive and well studied class of systematic pragmatic enrichments are scalar
implicatures
\citep{horn:1984,Hirschberg1985:A-Theory-of-Sca,LevinsonPragmatics1983,Geurts2010:Quantity-Implic}. Usually,
the utterance of a sentence like ``I own some of Johnny Cash's albums'' will be taken to mean
that the speaker does not own all of them. This is because, if the speaker instead owned them all, she
could have used the word \emph{all} instead of \emph{some} in her utterance, thereby making a
more informative statement. Scalar implicatures, especially the inference from \emph{some} to
\emph{some but not all}, have been studied extensively, both theoretically
\citep[e.g.,][]{Sauerland2004:Scalar-Implicat,ChierchiaFox2008:The-Grammatical,Rooyvan-RooijJagerde-Jager2012:Explaining-Quan}
as well as experimentally
\citep[e.g.,][]{BottNoveck2004:Some-Utterances,huang+snedeker:2009,GrodnerKlein2010:Some-and-Possib,GoodmanStuhlmuller2013:Knowledge-and-I,DegenTanenhaus2012:Processing-Scal}. This makes them particularly suitable candidates for the study of the evolution of regular pragmatic inferences.  While there is much dispute in this domain about many details, a position endorsed by a clear majority is that a scalar item like \emph{some} is underspecified to mean \emph{some and maybe all} and that the enrichment to \emph{some but not all} is part of some regular process with
roots in pragmatics.

If this majority view is correct, the question arises how such a division of labor between
semantics and pragmatics could have evolved; why it would be so pervasive across natural
languages; and why it is that some expressions systematically draw from it while others semantically conventionalize.

Models of language evolution abound. There are simulation-based models studying
populations of communicating agents
\citep{Hurford1989:Biological-Evol,Steels1995:A-Self-Organizi,LenaertsJansen2005:The-Evolutionar,SteelsBelpaeme2005:Coordinating-Pe,BaronchelliPuglisi2008:Cultural-route-,steels:2011,SpikeStadler2016:Minimal-Require}
and there are mathematical models of language evolution, many coming from game theory
\citep{Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,nowak:2006,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}. Much
of this work has focused on explaining basic properties such as compositionality and
combinatoriality
\citep[e.g.,][]{Batali1998:Computational-S,nowak+krakauer:1999,nowak+etal:2000,KirbyHurford2002:The-Emergence-o,kirby:2002,SmithKirby2003:Iterated-Learni,Gong2007:Language-Evolut,kirby+etal:2015,verhoef+etal:2014,Franke2015:Proto-Syntax},
but little attention has been paid to the interaction between conventional meaning and
pragmatic use. What is more, many mathematical models explain evolved meaning as a regularity
in the overt behavior of agents, abstracting from complex interactions between semantic
representations and pragmatic use. In contrast, we will here look at language users with a
richer linguistic make-up. More precisely, we spell out a model of the evolution of conventional meaning and pragmatic reasoning.
The objects of replication and selection are pairs consisting of a set of lexical
meanings and a manner of pragmatic behavior, which together define agents' linguistic choices. Pragmatic behaviors are represented using probabilistic models of pragmatic language use \citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
and selection are described by the \emph{replicator mutator dynamic}, a general and established
model of evolutionary change in large and homogeneous populations
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,nowak:2006}. This
approach allows us to study the interaction between (i) pressure towards communicative
efficiency and (ii) infidelity in the transmission of linguistic knowledge, caused by factors
such as inductive learning biases and sparse learning data \citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,KirbyGriffith2014:Iterated-Learni}. Considering the effects that the transmission of
linguistic knowledge through (iterated) learning have is important because neither semantic meaning nor pragmatic usage
patterns are directly observable. Instead, language learners have to infer these unobservables from the observable behavior in which they result. We formalize this process as a form of
Bayesian inference. Our approach thereby contains a well-understood model of
iterated Bayesian learning \citep{griffiths+kalish:2005,griffiths+kalish:2007}, but combines it with functional
selection, here formalized as the most versatile dynamic from evolutionary game theory, the
replicator dynamic \citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces
this model.

Section~\ref{sec:si-case-study} applies this model to a case study on scalar implicatures. We
discuss a setting in which the majority view of underspecified lexical meanings and systematic pragmatic
enrichments emerges if selection and transmission infidelity are combined. In particular, we
show that inductive learning biases of Bayesian learners that favor simpler lexical meanings
may play a role in preventing the lexicalization of scalar inferences and thereby lead to the emergence of Gricean-like
pragmatic reasoning types. The results of this study are critically assessed in the light of
the assumptions that feed our model in Section~\ref{sec:discussion}.

@@

\subsection{The replicator mutator dynamic}
An elegant formal approach to capture the interaction between selection gor communicative
efficiency and learnability is the \emph{replicator mutator dynamic}
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,nowak:2006}. The
RMD is a population dynamic. It describes the average change in the composition of a community
of agents of different types. The dynamic assumes a virtually infinite population, so that it
is able to abstract away from what individual agents do. The result is an abstract description
of the most probable trajectory of change in the population vector $\vec{x}$, which codifies
relative frequencies of types, so that $x_i$ is the relative frequency of type $i$ and
$\sum\nolimits_i x_i = 1$. In its simplest discrete-time formulation, the RMD defines the
frequency $x'_i$ of each type $i$ in the population at the next time step as a function of: (i)
the frequency $x_i$ of each type $i$ before the update step, (ii) the fitness $f_i$ of each
type $i$ before the update, and (iii) the probability $Q_{ji}$ that an agent who observes the
behavior of an agent with type $j$ ends up acquiring type $i$:
\begin{align}
  \label{eq:RMD_discrete}
  x'_i = \sum_j \frac{x_jf_j Q_{ji}}{\Phi}\,,
\end{align}
where $\Phi = \sum_k x_k f_k$ is the average fitness in the population.

The RMD consists of two components: fitness-based selection and transmission biases encoded in
the so-called mutation matrix $Q$. This becomes most transparent when we consider an equivalent
formulation in terms of a step-wise application of the discrete-time replicator dynamic
\citep{TaylorJonker1978:Evolutionary-St} on the initial population vector $\vec{x}$ and a
subsequent multiplication with the mutation matrix $Q$:
\begin{align}
  \label{eq:RMD_discrete_recast}
  x'_i & = (\text{M}(\text{RD}(\vec{x})))_i\,,
\end{align}
where
\begin{align*}
      \left ( \text{RD}(\vec{x}) \right )_i 
         = \frac{x_i f_i}{\Phi}
 \ \ \ \ \text{and} \ \ \ \ 
  (\text{M}(\vec{x}))_i = (\vec{x} \cdot Q)_i = \left ( \sum_j
          x_j Q_{ji} \right)_i\,.
\end{align*}
%
If there is no mutation, so that types are always replicated faithfully with $Q_{ji}=1$
whenever $j=i$, the RMD reduces to the replicator dynamic. If fitness is the same for all
types, $f_i = f_j$ for all $i$ and $j$, the RMD reduces to a process of iterated
mutation. Appendix~\ref{sec:minim-example-repl} provides an abstract minimal example of how the
replicator mutator dynamic works.

@@

The RMD is an abstract description of population-level changes. It abstracts away from specific
assumptions about how agents interact. To understand this better, the following looks at each
component in isolation: first replication, then mutation.

According to the replicator dynamic the relative frequency $x_i$ of type $i$ increases
proportional to its average fitness in the population. This dynamic is popular and versatile
because it can be derived from many processes of biological and cultural transmission and
selection \citep[for overview and several derivations
see][]{Sandholm2010:Population-Game}. Under a biological interpretation, the model describes
population-level changes when the number of offspring under single-parent preproduction is
proportional to the parent's fitness. Under a cultural interpretation, examples of concrete
agent-level processes that lead to population-level changes as described by the replicator
dynamic include conditional imitation
\citep[e.g.,][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and} or reinforcement
learning \citep[e.g.,][]{BorgersSarin997:Learning-Throug,Beggs2005:On-the-Converge}. In
conditional imitation, for instance, each agent carries a fixed type $i$ which it may change to
a type $j$ when it happens to observe another agent using type $j$. Imitation occurs with a
probability that is a monotonic function of how well $j$ fares against the population: the
better the behavior $j$, the more likely it is to be imitated.\footnote{A crucial assumption in
  all of the derivations of the replicator dynamic as a biological or cultural process of
  evolution is that agents themselves are rather limited in their strategic capacities. They
  are, for example, not able to innovate a strategy, i.e., to just choose a type, that is
  optimal given the current population state. They are also not able to strategically adapt
  their behavior to any particular agent they are interacting with on some given occasion. Such
  minimalism about the strategic capacities of agents is a conservative methodological
  choice. It allows us to study ---as the formal thought experiment that any such model is---
  how fitness-based selection changes a population in the absence of optimal agent-level
  reasoning and it abstracts away from specific assumptions about how fitness-based selection
  might be realized concretely.}

@@

\subsubsection{Types: Lexica and pragmatic strategies}
\label{sec:languages+use}
Types are what evolution operates on. They define an agent's fitness, usually through a payoff
accrued in single interactions with other agents. Often types can be identified as the possible
acts in a game, e.g., either cooperating or defecting in a prisoner's dilemma. In other cases,
they may be thought of as general properties of an agent that influence her fitness, such as
being positive or negative in our previous example (whatever that means). Since we are
interested in the evolutionary competition between different lexical representations and ways
of deploying them in communication, a type is here defined as a pair consisting of a lexicon
and a linguistic strategy of language use.

Agents play signaling games, in which the speaker wants to communicate a world state $s$ with a
message $m$ to a hearer who receives $m$ but does not know $s$
\citep[e.g.][]{lewis:1969,Skyrms2010:Signals}. A lexicon associates each message with a
(possibly fuzzy) set of states. A linguistic behavior specifies a probabilistic speaker rule (a
probabilistic choice of message for each state) and a probabilistic hearer rule (a
probabilistic choice of state for each message) given a lexicon. There are many ways of making
these general notions concrete. Since our main emphasis is on how different kinds of language
can be embedded within the replicator mutator dynamic, we focus here on a general formulation
of speaker and listener behavior which is as simple as possible but still in line with both a
game-theoretic tradition
\citep[e.g.,][]{benz:2006,benz+vRooij:2007,jaeger:2007b,franke+jaeger:2014} and a Bayesian
tradition
\citep[e.g.,][]{frank+goodman:2012,goodman+stuhlmueller:2013,GoodmanFrank2016:Pragmatic-Langu}. Here
is what we will assume in the remainder of this paper.

Lexica codify the truth-conditions of expressions. A convenient way to represent lexica is by
$(\card{\States}, \card{\Messgs})$-Boolean matrices, where $\States$ is a set of states
(meanings) and $M$ a set of messages (forms available in the language). For example, suppose
that there are two relevant world states $\States = \set{\ssome, \sall}$. In state $\ssome$
Chris owns some but not all of Johnny Cash's albums while in $\sall$ Chris owns them
all. Suppose that there are two messages $\Messgs = \set{\msome, \mall}$ where $\msome$ is
short for a sentence like \emph{Chris owns some of Johnny Cash's albums} and $\mall$ for the
same sentence with \emph{some} replaced by \emph{all}.  Lexica for this case would assign a
Boolean truth value, either $0$ for false or $1$ for true, to each state-message pair. The
following two lexica are minimal examples for the distinction between a lexicalized upper-bound
for \emph{some}, in $\Lbound$, and the widely assumed logical semantics with only a lower-bound,
in $\Llack$.

\begin{align*}
  \Lbound & = \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & 1 & 0 \\
      \mygray{\sall}  & 0 & 1 \\
    \end{block}
  \end{blockarray} &
  \Llack & = \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & 1 & 0 \\
      \mygray{\sall}  & 1 & 1 \\
    \end{block}
  \end{blockarray}
\end{align*}

@@

According to (\ref{h:level0}), a literal hearer's interpretation of a message depends on her
lexicon and her prior over states, $pr \in \Delta(S)$, which is here assumed flat for
simplicity. Literal interpreters thereby choose an arbitrary true interpretation for each
message according to their lexicon. Pragmatic hearers, defined in (\ref{h:leveln}), instead use Bayes' rule to weigh
interpretations based on a conjecture about speaker behavior. Speaker behavior is regulated by
a soft-max parameter $\lambda \geq 0$ \citep{luce:1959,sutton+barto:1998}. As
$\lambda$ increases, choices approximate strict maximization of expected utilities. Expected
utility of a message $\messg$ in state $\state$ for a level $n+1$ speaker is here defined as
$H_{n}(s|m;L)$, the probability that the hearer will assign to or choose the correct
meaning. For literal speakers, utility only tracks truthfulness. Literal speakers choose any
true message with equal probability but may send false messages as well with a probability
dependent on $\lambda$.

@@

There are two particularly important things to note. First, in contrast to their literal counterparts of level $0$, pragmatic agents of level $1$ using $\Llack$ associate $\msome$ preferentially with $\ssome$. This association is not perfect, and usually less strong than what agents with a lexicalized upper bound in $\Lbound$ can achieve -- with or without pragmatic reasoning. Higher order reasoning beyond level $1$ leads to stronger associations of $\msome$ and $\ssome$ also for the receiver. Still, the case study presented in Section~\ref{sec:si-case-study} will consider
sender and receiver behavior at levels 0 and 1, as the latter are the simplest pragmatic reasoning types which show a tendency to communicatively attuned pragmatic enrichment. Using only level 1 reasoning and possibly small $\lambda$ is therefore a conservative choice that works against the fitness-based selection of pragmatic language use for a notion of fitness defined as communicative success, which is introduced next. Second, when it comes to competition between types of use of lexicon $\Lbound$, pragmatic reasoning at level $1$ is not advantageous. The reason for this is that literal use of $\Lbound$ already endows agents with a behavioral strategy that associates a single state with a single message (in tendency; depending on $\lambda$ for senders). For $\Lbound$-receivers of level $1$, reasoning over stochasticity introduced at $S_0( \cdot \mid \cdot, \Lbound)$ will generally decrease the association of one state with one message. This decrease is only slight if $\lambda$ is high, but nevertheless present. That is to say, level-$1$ reasoning does not necessarily confer a functional advantage. For some types, such as users of $\Lbound$, literal signaling is preferable.

@@

n the context of language evolution, fitness is usually associated with the ability to
successfully communicate with other language users from the same population
\citep[e.g.,][]{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}. Concretely, the fitness
of type $i$ is its average expected communicative success, or \emph{expected utility} (EU),
given the relative frequencies $x_j$ of types $j$ in the current population:
\begin{align*}
  f_i = \sum_j x_j \text{EU}(t_i,t_j)\,.
\end{align*}
In signaling games, the expected utility $\text{EU}(t_i,t_j)$ for type $i$ when communicating with type $j$ is the
average success of $i$ when talking or listening to $j$. If, as standardly assumed, agents are speakers half of the
time, this yields:
\begin{align*}
  \text{EU}(t_i,t_j) = \nicefrac{1}{2} \, \text{EU}_S(t_i,t_j) + \nicefrac{1}{2} \, \text{EU}_H(t_i,t_j)\,,
\end{align*}
where $\text{EU}_S(t_i,t_j)$ and $\text{EU}_H(t_i,t_j)$ are the expected utilities for $i$ as a
speaker and as a hearer when communicating with $j$, defined as follows, where $n_i$ and $n_j$
are type $i$'s and type $j$'s pragmatic reasoning types and $L_i$ and $L_j$ are their lexica:
\begin{flalign*}
  & \text{EU}_S(t_i,t_j)  = \sum_s P(s)\sum_m S_{n_i}(m \mid s;L_i) \sum_{s'} H_{n_j}(s' \mid m;L_j)
  \delta(s,s') \,, \\
 & \text{EU}_H(t_i,t_j)  = \text{EU}_S(t_j,t_i)\,.
\end{flalign*}
As usual, we assume that agents are cooperative: $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.

>>>>

\begin{abstract}
  According to standard linguistic theory, the meaning of an utterance is the product of
  conventional semantic meaning and general pragmatic rules on language use. To investigate how
  cultural evolution of language plays out under this picture of the semantics-pragmatics
  interface, we present a game-theoretic model of the competition between types of language
  users, each endowed with certain lexical representations and a particular pragmatic
  disposition to act on them. Our model traces two evolutionary forces and their interaction:
  (i) pressure towards communicative efficiency and (ii) transmission perturbations during the
  acquisition of linguistic knowledge. We illustrate the model based on a case study on scalar
  implicatures, which suggests that the relationship between underspecified semantics and
  pragmatic inference is one of co-evolution. The evolution of underspecified semantics
  requires pragmatic reasoning, while pragmatic reasoning may not evolve if lexical meanings do
  not require pragmatic enrichment.
\end{abstract}

@@

A particularly productive and well studied class of systematic pragmatic enrichments are scalar
implicatures
\citep{horn:1984,Hirschberg1985:A-Theory-of-Sca,LevinsonPragmatics1983,Geurts2010:Quantity-Implic}. Usually,
the utterance of a sentence like ``I own some of Johnny Cash's albums'' will be taken to mean
that the speaker does not own all of them. This is because, if the speaker instead owned them
all, she could have used the word \emph{all} instead of \emph{some} in her utterance, thereby
making a more informative statement. Scalar implicatures, especially the inference from
\emph{some} to \emph{some but not all}, have been studied extensively, both theoretically
\citep[e.g.,][]{Sauerland2004:Scalar-Implicat,ChierchiaFox2008:The-Grammatical,Rooyvan-RooijJagerde-Jager2012:Explaining-Quan}
as well as experimentally
\citep[e.g.,][]{BottNoveck2004:Some-Utterances,huang+snedeker:2009,GrodnerKlein2010:Some-and-Possib,GoodmanStuhlmuller2013:Knowledge-and-I,DegenTanenhaus2012:Processing-Scal}. This
makes them particularly suitable candidates for the study of the evolution of regular pragmatic
inferences.  While there is much dispute in this domain about many details, a position endorsed
by a clear majority is that a scalar item like \emph{some} is underspecified to mean \emph{some
  and maybe all} and that the enrichment to \emph{some but not all} is part of some regular
process with roots in pragmatics. If this majority view is correct, the question arises how
such a division of labor between semantics and pragmatics could have evolved.

Models of language evolution abound. There are simulation-based models studying populations of
communicating agents
\citep{Hurford1989:Biological-Evol,Steels1995:A-Self-Organizi,LenaertsJansen2005:The-Evolutionar,SteelsBelpaeme2005:Coordinating-Pe,BaronchelliPuglisi2008:Cultural-route-,steels:2011,SpikeStadler2016:Minimal-Require}
and there are mathematical models of language evolution, many coming from game theory
\citep{Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,nowak:2006,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}. Much
of this work has focused on explaining basic properties such as compositionality and
combinatoriality
\citep[e.g.,][]{Batali1998:Computational-S,nowak+krakauer:1999,nowak+etal:2000,KirbyHurford2002:The-Emergence-o,kirby:2002,SmithKirby2003:Iterated-Learni,Gong2007:Language-Evolut,kirby+etal:2015,verhoef+etal:2014,Franke2015:Proto-Syntax},
but little attention has been paid to the interaction between conventional meaning and
pragmatic use. What is more, many mathematical models explain evolved meaning as a regularity
in the overt behavior of agents, abstracting from complex interactions between semantic
representations and pragmatic use.

In contrast, we will here look at language users with a richer linguistic make-up. The objects
of replication and selection in our model are pairs consisting of a set of lexical meanings and
a manner of pragmatic behavior, which together define an agent's linguistic choices. Types of
pragmatic behavior are represented using probabilistic models of pragmatic language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
and selection are described by the \emph{replicator mutator dynamic}, a general and established
model of evolutionary change in large and homogeneous populations
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,nowak:2006}. This
approach allows us to study the interaction between (i) pressure towards communicative
efficiency and (ii) infidelity in the transmission of linguistic knowledge, caused by factors
such as inductive learning biases and sparse learning data
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,KirbyGriffith2014:Iterated-Learni}. Considering
the effects that the transmission of linguistic knowledge through (iterated) learning have is
important because neither semantic meaning nor pragmatic usage patterns are directly
observable. Instead, language learners have to infer these unobservables from the observable
behavior in which they result. We formalize this process as a form of Bayesian inference. Our
approach thereby contains a well-understood model of iterated Bayesian learning
\citep{griffiths+kalish:2005,griffiths+kalish:2007}, but combines it with functional selection,
here formalized as the most versatile dynamic from evolutionary game theory, the replicator
dynamic \citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces this
model.

Section~\ref{sec:si-case-study} applies this model to a case study on scalar implicatures. It
shows that inductive learning biases of Bayesian learners that favor simpler lexical meanings
may play a role in preventing the lexicalization of scalar inferences and thereby lead to the
emergence of Gricean-like pragmatic reasoning types, but only if there is selective pressure
for communicative efficiency and learnability. The results of this study are critically
assessed in the light of the assumptions that feed our model in Section~\ref{sec:discussion}.

@@

\subsection{The replicator mutator dynamic}
An elegant formal approach to capture the interaction between selection for communicative
efficiency and learnability is the \emph{replicator mutator dynamic}
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,nowak:2006}. The
RMD is a population dynamic which describes the average change in the composition of a
community of agents of different types. The dynamic assumes a virtually infinite population, so
that it is able to abstract away from what individual agents do. It describes the most probable
trajectory of change in the population vector $\vec{x}$, which codifies relative frequencies of
types, so that $x_i$ is the relative frequency of type $i$ and $\sum\nolimits_i x_i = 1$. In
its simplest discrete-time formulation, the RMD defines the frequency $x'_i$ of each type $i$
in the population at the next time step as a function of: (i) the frequency $x_i$ of each type
$i$ before the update step, (ii) the fitness $f_i$ of each type $i$ before the update, and
(iii) the probability $Q_{ji}$ that an agent who observes the behavior of an agent with type
$j$ ends up acquiring type $i$ (with $\sum_k Q_{jk} = 1$ for all $j$):
\begin{align}
  \label{eq:RMD_discrete}
  x'_i = \sum_j \frac{x_jf_j Q_{ji}}{\Phi}\,,
\end{align}
where $\Phi = \sum_k x_k f_k$ is the average fitness in the population.

The RMD consists of two components: fitness-based selection and transmission biases encoded in
the so-called mutation matrix $Q$. This becomes most transparent when we consider an equivalent
formulation in terms of a step-wise application of the discrete-time replicator dynamic
\citep{TaylorJonker1978:Evolutionary-St} on the initial population vector $\vec{x}$ and a
subsequent multiplication with the mutation matrix $Q$:
\begin{align}
  \label{eq:RMD_discrete_recast}
  x'_i & = (\text{M}(\text{RD}(\vec{x})))_i\,,
\end{align}
where
\begin{align*}
      \left ( \text{RD}(\vec{x}) \right )_i 
         = \frac{x_i f_i}{\Phi}
 \ \ \ \ \text{and} \ \ \ \ 
  (\text{M}(\vec{x}))_i = (\vec{x} \cdot Q)_i = \left ( \sum_j
          x_j Q_{ji} \right)_i\,.
\end{align*}
%
If there is no mutation, so that types are always replicated faithfully with $Q_{ji}=1$
whenever $j=i$, the RMD reduces to the replicator dynamic. If fitness is the same for all
types, $f_i = f_j$ for all $i$ and $j$, the RMD reduces to a process of iterated
mutation. Appendix~\ref{sec:minim-example-repl} provides an abstract minimal example.

@@

The RMD is an abstract description of population-level changes. It abstracts away from specific
assumptions about how agents interact and what causes selection and mutation. To understand
this better, the following looks at each component in isolation: first replication, then
mutation.

According to the replicator dynamic the relative frequency $x_i$ of type $i$ increases
proportional to its average fitness in the population. This dynamic is popular and versatile
because it can be derived from many processes of biological and cultural transmission and
selection \citep[for overview and several derivations
see][]{Sandholm2010:Population-Game}. Under a biological interpretation, the model describes
population-level changes when the number of offspring under single-parent reproduction is
proportional to the parent's fitness. Under a cultural interpretation, examples of concrete
agent-level processes that lead to population-level changes as described by the replicator
dynamic include conditional imitation
\citep[e.g.,][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and} or reinforcement
learning \citep[e.g.,][]{BorgersSarin997:Learning-Throug,Beggs2005:On-the-Converge}. In
conditional imitation, for instance, each agent carries a fixed type $i$ which it may change to
a type $j$ when it happens to observe another agent using type $j$. Imitation occurs with a
probability that is a monotonic function of $f_j$, i.e., of how well $j$ fares against the
population: the better the behavior $j$, the more likely it is to be imitated.\footnote{A
  crucial assumption in all of the derivations of the replicator dynamic as a biological or
  cultural process of evolution is that agents themselves are rather limited in their strategic
  capacities. They are, for example, not able to innovate a strategy, i.e., to just choose a
  type, that is optimal given the current population state. They are also not able to
  strategically adapt their behavior to any particular agent they are interacting with on some
  given occasion. Such minimalism about the strategic capacities of agents is a conservative
  methodological choice. It allows us to study ---as the formal thought experiment that any
  such model is--- how fitness-based selection changes a population in the absence of optimal
  agent-level reasoning and it abstracts away from specific assumptions about how fitness-based
  selection might be realized concretely.}

@@

\subsubsection{Types: Lexica and pragmatic strategies}
\label{sec:languages+use}
Types are what evolution operates on. A type is what determines an agent's behavior and thereby
an agent's fitness, so that it makes sense to speak of the fitness of a type itself. To study
the joint evolution of semantic meaning and pragmatic use, types are pairs consisting of a
lexicon and a linguistic strategy of language use.

Agents play signaling games, in which the speaker wants to communicate a world state $s$ with a
message $m$ to a hearer who receives $m$ but does not know $s$
\citep[e.g.][]{lewis:1969,Skyrms2010:Signals}. A lexicon associates each message with a set of
states. A linguistic behavior specifies a probabilistic speaker rule (a probabilistic choice of
message for each state) and a probabilistic hearer rule (a probabilistic choice of state for
each message) given a lexicon. There are many ways of making these general notions
concrete. In the following we aim at a general formulation of speaker and listener
behavior which is as simple as possible but still in line with both a game-theoretic tradition
\citep[e.g.,][]{benz:2006,benz+vRooij:2007,jaeger:2007b,franke+jaeger:2014} and a Bayesian
tradition
\citep[e.g.,][]{frank+goodman:2012,goodman+stuhlmueller:2013,GoodmanFrank2016:Pragmatic-Langu}.

Lexica codify the truth-conditions of expressions. A convenient way to represent lexica is by
$(\card{\States}, \card{\Messgs})$-Boolean matrices, where $\States$ is a set of states
(meanings) and $M$ a set of messages (forms available in the language). For example, suppose
that there are two relevant world states $\States = \set{\ssome, \sall}$. In state $\ssome$
Chris owns some but not all of Johnny Cash's albums while in $\sall$ Chris owns them
all. Suppose that there are two messages $\Messgs = \set{\msome, \mall}$ where $\msome$ is
short for a sentence like \emph{Chris owns some of Johnny Cash's albums} and $\mall$ for the
same sentence with \emph{some} replaced by \emph{all}.  Lexica then assign a Boolean truth
value, either $0$ for false or $1$ for true, to each state-message pair. The following two
lexica are minimal examples for the distinction between a lexicalized upper-bound for
\emph{some}, in $\Lbound$, and the widely assumed logical semantics with only a lower-bound, in
$\Llack$.
\begin{align*}
  \Lbound & = \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & 1 & 0 \\
      \mygray{\sall}  & 0 & 1 \\
    \end{block}
  \end{blockarray} &
  \Llack & = \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & 1 & 0 \\
      \mygray{\sall}  & 1 & 1 \\
    \end{block}
  \end{blockarray}
\end{align*}

@@

According to (\ref{h:level0}), a literal hearer's interpretation of a message depends on her
lexicon and her prior over states, $pr \in \Delta(S)$, which is here assumed flat for
simplicity. Literal interpreters thereby choose an arbitrary true interpretation for each
message according to their lexicon. Pragmatic hearers, defined in (\ref{h:leveln}), instead use
Bayes' rule to weigh interpretations based on a conjecture about speaker behavior. Speaker
behavior is regulated by a soft-max parameter $\lambda \geq 0$
\citep{luce:1959,sutton+barto:1998}. As $\lambda$ increases, choices approximate strict
maximization of expected utilities. Expected utility of a message $\messg$ in state $\state$
for a level $n+1$ speaker is here defined as $H_{n}(s|m;L)$, the probability that the hearer
will assign to or choose the correct meaning.\footnote{An alternative choice for the speaker's
  utility function is $\log H_{n}(s|m;L)$. Conceptually, the difference is that without the
  logarithm, utilities are derived from action choices, while with the logarithm utilities are
  derived from how closely the listener's beliefs (after hearing an utterance) align with the
  speaker's, as measure by Kullback-Leibler divergence
  \cite{frank+goodman:2012,qing+franke:2015}. In an evolutionary context, where fitness comes
  from actual interactions with the environment, not just beliefs held about it, the
  formulation without the logarithm may be slightly more appealing. Moreover, log-based speaker
  utilities cause the speaker never to use false utterances. Since we aim here for a situation
  in which any type has a reasonable chance to be mistaken for any other based on their
  production behavior (see Section~\ref{sec:learnability}), we want a definition of speaker
  types which also send any message in any state with positive probability.} For literal
speakers, utility only tracks truthfulness. Literal speakers choose any true message with equal
probability but may send false messages as well with a probability dependent on $\lambda$.

@@

In contrast to their literal counterparts of level $0$, pragmatic agents of level $1$ who use
$\Llack$ associate $\msome$ preferentially with $\ssome$. This association is not perfect, and
usually less strong than what agents with a lexicalized upper bound in $\Lbound$ can achieve --
with or without pragmatic reasoning. Higher order reasoning beyond level $1$ leads to stronger
associations of $\msome$ and $\ssome$ also for the receiver. Still, the case study presented in
Section~\ref{sec:si-case-study} will consider sender and receiver behavior at levels 0 and 1,
as the latter are the simplest pragmatic reasoning types which show a tendency to
communicatively attuned pragmatic enrichment. Using only level 1 reasoning and possibly small
$\lambda$ is therefore a conservative choice that works \emph{against} the fitness-based
selection of pragmatic language use for a notion of fitness defined as communicative success,
which is introduced next.

When it comes to competition between types of use of lexicon $\Lbound$, pragmatic reasoning at
level $1$ is not advantageous. The reason for this is that literal use of $\Lbound$ already
endows agents with a behavioral strategy that associates a single state with a single message
(in tendency; depending on $\lambda$ for senders). For $\Lbound$-receivers of level $1$,
reasoning over stochasticity introduced at $S_0( \cdot \mid \cdot, \Lbound)$ will generally
decrease the association of one state with one message. This decrease is only slight if
$\lambda$ is high, but nevertheless present. That is to say, level-$1$ reasoning does not
necessarily confer a functional advantage. For some types, such as users of $\Lbound$, literal
signaling is preferable. In sum, the definition of types introduced here constitute conditions
that are purposefully averse to what we would like to show; the evolution of is not immediate
just based on fitness-based selection, to which we turn next.

@@

In the context of language evolution, fitness is usually associated with the ability to
successfully communicate with other language users from the same population
\citep[e.g.,][]{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}. Concretely, the fitness
of type $i$ is its average expected communicative success, or \emph{expected utility} (EU),
given the relative frequencies $x_j$ of types $j$ in the current population:
\begin{align*}
  f_i = \sum_j x_j \text{EU}(t_i,t_j)\,.
\end{align*}
The expected utility $\text{EU}(t_i,t_j)$ for type $i$ when communicating with type $j$ is the
average success of $i$ when talking or listening to $j$. If, as standardly assumed, agents are speakers half of the
time, this yields:
\begin{align*}
  \text{EU}(t_i,t_j) = \nicefrac{1}{2} \, \text{EU}_S(t_i,t_j) + \nicefrac{1}{2} \, \text{EU}_H(t_i,t_j)\,,
\end{align*}
where $\text{EU}_S(t_i,t_j)$ and $\text{EU}_H(t_i,t_j)$ are the expected utilities for $i$ as a
speaker and as a hearer when communicating with $j$, defined as follows, where $n_i$ and $n_j$
are type $i$'s and type $j$'s pragmatic reasoning types and $L_i$ and $L_j$ are their lexica:
\begin{flalign*}
  & \text{EU}_S(t_i,t_j)  = \sum_s P(s)\sum_m S_{n_i}(m \mid s;L_i) \sum_{s'} H_{n_j}(s' \mid m;L_j)
  \delta(s,s') \,, \\
 & \text{EU}_H(t_i,t_j)  = \text{EU}_S(t_j,t_i)\,.
\end{flalign*}
As usual, we assume that agents are cooperative: $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.
